%% Paper implementing score test framework for SCR :)

%% Uncomment to knit for Ben and Rachel:
\documentclass[14pt]{extarticle}
\usepackage{geometry}
\geometry{left=20mm, right=20mm, top=5mm, bottom=5mm,
 headheight=5mm, marginpar=5mm}

%% Uncomment to knit for writing: 
% \documentclass{article}
% \usepackage{geometry}
% \geometry{left=15mm, right=15mm, top=5mm, bottom=5mm,
%  headheight=5mm, marginpar=5mm}

\usepackage{layout}
\usepackage{blindtext}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{multicol}
%\usepackage{subfig}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage[parfill]{parskip} % Want line break between paragraphs instead of indentation
\usepackage{setspace} % For line spacing
\usepackage{lineno,xcolor} % For numbering lines

 % For R code, for now :)
\usepackage{listings}
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
\lstdefinestyle{mystyle}{
  backgroundcolor=\color{backcolour},   
  commentstyle=\color{codegreen},
  basicstyle=\ttfamily\footnotesize,
  breakatwhitespace=false,         
  breaklines=true,                 
  captionpos=b,                    
  keepspaces=true,                 
  numbers=left,                    
  numbersep=5pt,                  
  showspaces=false,                
  showstringspaces=false,
  showtabs=false,                  
  tabsize=2
}
\lstset{style=mystyle}


%% Bibliography, referencing 
\usepackage[backend=biber,citestyle=authoryear,uniquename=false,maxcitenames=2,
mincitenames=1,uniquelist=false, style=apa,giveninits=true]{biblatex}
\addbibresource{../../../../../PhDReferences_Zotero.bib}
\makeatletter
\newcommand{\apamaxcitenames}{2}
\DeclareNameFormat{labelname}{%
  % First set the truncation point
  \ifthenelse{\value{uniquelist}>1}
    {\numdef\cbx@min{\value{uniquelist}}}
    {\numdef\cbx@min{\value{minnames}}}%
  % Always print the first name and the second if there are only two since
  % "et al" must always be plural
  \ifboolexpr{test {\ifnumcomp{\value{listcount}}{=}{1}}
              or test {\ifnumcomp{\value{listtotal}}{=}{2}}}
    {\usebibmacro{labelname:doname}%
      {\namepartfamily}%
      {\namepartfamilyi}%
      {\namepartgiven}%
      {\namepartgiveni}%
      {\namepartprefix}%
      {\namepartprefixi}%
      {\namepartsuffix}%
      {\namepartsuffixi}}
    % We are looking at name >=3
    % If the list is 6 or more names or we have seen citation before, potential truncation
    {\ifboolexpr{test {\ifnumcomp{\value{listtotal}}{>}{\apamaxcitenames}}
                 or test {\ifciteseen}}
     % Less than the truncation point, print normally
     {\ifnumcomp{\value{listcount}}{<}{\cbx@min + 1}
       {\usebibmacro{labelname:doname}%
         {\namepartfamily}%
         {\namepartfamilyi}%
         {\namepartgiven}%
         {\namepartgiveni}%
         {\namepartprefix}%
         {\namepartprefixi}%
         {\namepartsuffix}%
         {\namepartsuffixi}}
       {}%
      % At potential truncation point ...
      \ifnumcomp{\value{listcount}}{=}{\cbx@min + 1}
        % but enforce plurality of et al - only truncate here if there is at
        % least one more element after the current potential truncation point
        % so that "et al" covers at least two elements.
        {\ifnumcomp{\value{listcount}}{<}{\value{listtotal}}
          {\printdelim{andothersdelim}\bibstring{andothers}}
          {\usebibmacro{labelname:doname}%
            {\namepartfamily}%
            {\namepartfamilyi}%
            {\namepartgiven}%
            {\namepartgiveni}%
            {\namepartprefix}%
            {\namepartprefixi}%
            {\namepartsuffix}%
            {\namepartsuffixi}}}
        {}%
      % After truncation point, do not print name
      \ifnumcomp{\value{listcount}}{>}{\cbx@min + 1}
       {\relax}%
       {}}%
     % We are looking at name >=3
     % Name list is < 6 names or we haven't seen this citation before, print normally
     {\usebibmacro{labelname:doname}%
       {\namepartfamily}%
       {\namepartfamilyi}%
       {\namepartgiven}%
       {\namepartgiveni}%
       {\namepartprefix}%
       {\namepartprefixi}%
       {\namepartsuffix}%
       {\namepartsuffixi}}}}
\makeatother

%% Notation stuff
\newcommand{\thetaZero}{\boldsymbol{\theta}_0}
\newcommand{\thetaZeroMLE}{\hat{\boldsymbol{\theta}}_0}
\newcommand{\thetaOne}{\boldsymbol{\theta}_1}
\newcommand{\thetaOneTilde}{\tilde{\boldsymbol{\theta}}_1}
\newcommand{\thetaOneMLE}{\hat{\boldsymbol{\theta}}_1}


\title{Harnessing score tests for model selection}

\begin{document}

\maketitle

% Setting double spacing
\doublespacing
% Adding line numbering
\linenumbers

\section{Introduction}
\label{sec:introduction}

Model selection is an important step for
statistical analysis, allowing the practitioner to confirm the
reliability of their chosen model and any resulting
conclusions. Well-known model selection methods include comparing
models using likelihood ratio tests (LRTs), AIC values, and score 
tests. Comparing two nested models via LRTs and AIC values means 
 using maximised log-likelihood values to assess the quality of
each model, while applying a penalty for model
complexity. Conducting a LRT requires twice the difference of each
model's maximised log-likelihood. This is 
compared to a null distribution, where
the degrees of freedom can be thought of as containing the penalty for
model complexity \parencite{vaartAsymptoticStatistics2012}. 
Using AIC values means directly adding a penalty term to each
maximised log-likelihood and comparing the results \parencite{burnham2002}. 

To compare two nested models using a score test, a score test
statistic is calculated and compared to the same null
distribution as the LRT statistic. However, the score test statistic
employs gradients to inform the improvement in fit available by
perturbing the parameter estimates from the simpler model
\parencite{rao2005}. Specifically, the score test statistic is
calculated using the
score vector and information matrix for the more complicated model,
evaluated at the parameter estimates of the simpler model. See
\textcite{rao2005} for an overview and recent developments of score
tests. 

Each method has its own advantages and
disadvantages. It is straightforward to perform the
calculations required to compare models via LRTs or AIC values. However, both
models must be fitted. To conduct score tests, only the simpler of the
two models needs to be fitted \parencite{rao2005}. However, the required
gradients can be challenging to obtain, often acting as a
deterrent for conducting these tests. 

Historically, LRTs and AIC values have been widely-used methods
of model selection. However, when the models under comparison require
considerable time and resources to fit, these methods of comparison can
become inconvenient. This issue is amplified when considering
multiple potential models for a data set, as every model in the
candidate set must be fitted, potentially rendering model selection
impractical.

In such cases, turning to score tests would be
beneficial as fitting fewer models makes model selection more
practical. However, for usability, we require a straightforward way of
calculating the gradients required by the score test
statistic. Traditionally, options for obtaining these gradients
included symbolic differentiation and numeric approximations. The former
can quickly become unrealistic as model complexity grows, while the
latter often produces unstable and inaccurate
results \parencite{ADMBpaper}. In this paper, we demonstrate that
due to recent developments in statistical software, automatic
differentiation (AD) tools make previously
unfeasible tasks, such as model selection via score tests, viable in a
way previously unseen. Repurposing the model-fitting capabilities of AD tools allows
for convenient procurement of the necessary gradients, and 
therefore a newly straightforward implementation of score tests. 

Harnessing the chain rule of differentiation, AD tools produce exact
gradients of any given function \parencite{ADMBpaper}. Using these
tools only requires the practitioner to supply the objective function 
\parencite{TMBpaper}. For statistical models, these tools are 
mainly used for model-fitting or calculating 
the Laplace approximation. Utilising the side-effects of these
capabilities -- repurposing their ability to calculate
exact log-likelihood gradients -- now facilities a
convenient implementation of score tests.

This has implications for a wide range of models. For instance, for
random effects models, intractable
likelihoods involving potentially high 
dimensional integrals produce gradients with no closed form, resulting in
a host of challenges. Model fitting previously required tailored
software, even for relatively simple examples such as
generalised linear mixed models (GLMMs), although gradients remained
elusive \parencite{lme4-pkg}. Being 
commonly-used modern statistical models, \textcite{wang2022} present extensive derivations
to obtain GLMM gradients, enabling valuable yet previously
inaccessible tasks such as score tests.
However, there remain many random effects 
models for which such solutions and derivations don't exist. Random effects
models are just one illustration of the  
advancements provided by AD. For many classes of models, the
convenience of fitting models and obtaining gradients via AD tools
makes endeavours such as model selection via
score tests newly accessible.

To illustrate the benefits of score tests 
and AD tools, we will focus on the class of random effects models known as
spatial capture-recapture (SCR) models. Model selection is often difficult 
for SCR models. Likelihood equations for these models involve products of two-dimensional spatial
integrals \parencite{borchersEfford2008}. This makes model-fitting, and therefore
using AIC values to compare 
models, time consuming and resource intensive,
especially when many models are being considered. However, an absence
of suitable alternatives means comparison via AIC values is a common
and established method of SCR model selection (e.g. \cite{secr-aic-ex}). Due 
to the difficulties in symbolic differentiation or numerically 
approximating gradients, score tests have not generally been
used. Removing the barriers
to conducting score tests 
presents a new and potentially impactful solution to this issue. 

We will explore the newfound gains of conducting score tests using AD tools
via the framework introduced by \textcite{catchpoleMorgan1996}, and
later developed by \textcite{mccreaMorgan2011}. Following the
convention of organising the
candidate set into a tree structure, the framework provides an efficient way of
navigating through the set using score tests, 
amplifying the benefits of score tests for model selection. The issue
of choosing a model from a   
potentially vast candidate set of models of varying complexity is reduced to the
task of calculating score statistics and fitting a few, potentially
relatively simple, models. 

We present our findings using an illustrative example, using an SCR
data set of ovenbird captures \parencite{secr-pkg}. As described
below, similarities to the data set presented in
\textcite{borchersEfford2008} mean we have a range of possible
parameter dependencies, and therefore many potential
models to consider. We employ
the `RTMB' R package \parencite{rtmb-pkg}, a widely-used AD tool, to
implement score tests for SCR model selection. Compared to traditional
methods, we examine how our approach enhances the 
convenience and accessibility of model selection, while accommodating for
potentially common issues such as non-nested models.

Our proposed model selection approach is widely applicable, it does not only
apply to SCR models or random effects models. We present a general
method, allowing accessible model selection for many classes of
models across a vast range of fields. 

\section{Model selection using score tests}
\label{sec:model-sel}

\subsection{Calculating score test statistics with RTMB}
\label{sec:calculating-stats-rtmb}

We begin with a brief overview of conducting score tests, refer to texts such
as \textcite{rao2005} for detailed information. When
comparing two nested models 
using a score test, the null hypothesis 
states the simpler model is adequate. We will refer to this
 model as the `$H_0$ model', and the more complex model as the `$H_1$
model'. We label the parameter vector for the $H_0$ model as `$\thetaZero$',
and the parameter vector for the $H_1$ model as `$\thetaOne$'. To
conduct a score test:
\begin{enumerate}
\item Fit the $H_0$ model, extract the Maximum Likelihood Estimate
  (MLE) vector $\thetaZeroMLE$; 
\item Translate $\thetaZeroMLE$ into a $\thetaOne$ vector --
  i.e. identify a $\thetaOne$ vector that shares the constraints of
  the $H_0$ model and uses the estimates of $\thetaZeroMLE$. By
  definition, this vector  makes the $H_1$ model
  equivalent to the fitted $H_0$ model from Step 1. We
  label the resulting vector $\thetaOneTilde$;
\item Calculate the score test statistic:
  \begin{equation}
    \boldsymbol{u}_1 (\thetaOneTilde)^T \mathcal{I}_1
    (\thetaOneTilde)^{-1} \boldsymbol{u}_1 (\thetaOneTilde)
  \end{equation}
  where $\boldsymbol{u}_1 (\thetaOneTilde)$ is the score vector of
  the $H_1$ model evaluated at $\thetaOneTilde$, and $\mathcal{I}_1
  (\thetaOneTilde)$ is the information matrix
  evaluated at $\thetaOneTilde$. Note that we will focus on using the
  observed information matrix, see the Discussion for more details;
\item Compare this statistic to a Chi-squared distribution where
  the degrees of freedom is the difference in the number
  of parameters for the $H_0$ and $H_1$ model, producing a p-value to
  assess $H_0$. 
\end{enumerate}

By considering the gradient of the $H_1$ log-likelihood at
$\thetaOneTilde$ via $\boldsymbol{u}_1
(\thetaOneTilde)$, and the curvature via $\mathcal{I}_1
(\thetaOneTilde)$, the score test statistic summarises the
distance between $\thetaOneTilde$ and the unconstrained $H_1$
MLE vector. If this distance is deemed sufficiently large by the
Chi-squared distribution from Step 4, the
difference in fit between the $H_0$ and $H_1$ model is enough for the
$H_1$ model to be preferable, despite its added complexity. 

Previously, the obstacle in implementing score tests was obtaining the
score vector and information matrix to evaluate at
$\thetaOneTilde$. Using recent software advancements, we employ 
the `RTMB' R package -- a commonly-used AD tool -- to do
this. Translating the $H_1$ log-likelihood into 
an R function, it becomes straightforward to extract the score vector
and information matrix. Using AD to also fit
the $H_0$ model can provide time
savings, especially for sophisticated models. See Code Block 1 for a
simple example. Due to its  
simplicity, using AD to fit the $H_0$ model in this example does not
necessarily provide a time saving, but is done for illustrative purposes. 

\begin{lstlisting}[language=R]
# We are working with a similar linear regression model
# We assume that: Y_i ~ Normal(mu_i, 1), where i=1,...,20
# Our H0 model is: mu_i = beta0
# Our H1 model is: mu_i = beta0 + beta1 * x_i, where x is a vector of 20 covariate values

# In the code below, we assume that:
# (1) The response values are saved as an object labelled `y'
# (2) The covariate values are saved as an object labelled `x'
# (3) Appropriate starting values for optimisation for beta0 and beta1 are saved as objects labelled `beta0.start' and `beta1.start', respectively

# Defining the objective function (negative log-likelihood) for the H0 model
H0.nll = function(params) {
  beta0 = params$beta0
  sum(dnorm(y, mean=beta0, sd=1, log=TRUE)) * (-1)
}

# Defining the objective function (negative log-likelihood) for the H1 model
H1.nll = function(params) {
  beta0 = params$beta0
  beta1 = params$beta1
  sum(dnorm(y, beta0 + (beta1 * x), 1, log=TRUE)) * (-1)
}

# Creating the RTMB object for the H0 model
H0.object = MakeADFun(func=H0.nll, parameters=list(beta0=beta0.start))
# Fitting the H0 model
H0.fit = nlminb(start=H0.object$par, objective=H0.object$fn, gradient=H0.object$gradient)
# Extracting the H0 MLE
H0.MLE = H0.fit$par

# Creating the RTMB object for the H1 model, from which we can extract and evaluate gradients
H1.object = MakeADFun(func=H1.nll, parameters=list(beta0=beta0.start, beta1=beta1.start))

# Constructing an H1 vector that shares the constraints and MLEs of the H0 model
H0.MLE.for.H1 = list(beta0=H0.MLE["beta0"], beta1=0)
# Evaluating the H1 score vector and information matrix (we multiply by -1 below as we are working with the negative log-likelihood)
score.vec = H1.object$gr(H0.MLE.for.H1) * (-1)
obs.info.mat = H1.object$he(H0.MLE.for.H1)

# Calculating the score statistic
score.statistic = score.vec %*% solve(obs.info.mat) %*% t(score.vec)
# Extracting the p-value
p.value = pchisq(score.statistic, df=1, lower=F)
\end{lstlisting}


\subsection{Implementing the score test framework}
\label{sec:implementing-framework}

By organising the candidate set into a tree structure, the model
selection framework of \textcite{catchpoleMorgan1996}
defines an efficient path to the final
model, beginning by considering the simplest models. This is done
through the construction of `levels', and the use of score tests to
select the best model at each level.

As described in \textcite{mccreaMorgan2011}, 
the most basic model defines Level 0 of the candidate set, and Level 1
contains all models 
with one added parameter dependency. Score tests are used to compare all Level 1
models to the Level 0 model. As in 
Section \ref{sec:calculating-stats-rtmb}, we refer to the Level 0
model as the `$H_0$ model', and the Level 1 models as `$H_1$
models'. The $H_1$ model associated with the smallest significant
p-value defines the new null hypothesis -- that is, it becomes the new
$H_0$ model. Level 2 is then constructed, consisting of all models with one
additional parameter dependency relative to this model, and the
process is repeated. We stop when none of the score tests produce
significant p-values. At this point, the $H_0$ model becomes the
chosen model.  

By using the level structure, implementing this
framework on a 
comprehensive candidate set likely only requires a
few models be fitted, as only the $H_0$ models need to be fitted for
the score tests. Furthermore, the stepwise nature of the  
framework promotes convenient model-fitting -- the MLEs
from a given $H_0$ model may provide appropriate starting values for
the next $H_0$ model \parencite{mccreaMorgan2011}.

Harnessing the utility of score tests and convenience of 
AD, the framework of \textcite{catchpoleMorgan1996} requires fewer
resources and less time than traditional model selection
techniques, especially for a comprehensive candidate set. Fitting
all models under consideration is replaced by accessible
calculations and fitting a few models, enhancing the accessibility of
model selection. 

\subsection{Issues with nesting}
\label{sec:issues-nesting-general}

During model selection with score tests, obstacles can arise due to
non-nested models, or nested models with boundary issues. With
non-nested models, it is impossible 
to obtain the $H_0$ model from the $H_1$ model, i.e. no
$\thetaOneTilde$ vector exists, so a
score test statistic cannot be calculated. Boundary issues occur when values in
$\thetaOneTilde$ lie on the boundary of the parameter space. The null
Chi-squared distribution no longer holds, so any p-values are
unusable \parencite{vaartAsymptoticStatistics2012}.

In such cases, we propose separating the
candidate set into groups -- where models in each group fall into a
nested structure with no boundary issues -- and carrying out the
framework of \textcite{catchpoleMorgan1996} on each group, using AIC values to
make a final selection from the chosen models. See
below for an example of this.

This proposed extension is unlikely to
require any extra models be fitted, as each group's final model is likely 
fitted during model selection. We
are proposing employing a well-established method for model
comparison, allowing practitioners to take advantage of score tests
via the framework of \textcite{catchpoleMorgan1996} in the presence of
nesting issues.  


% \section{Application: ovenbirds}
% \label{sec:ovenbirds}

% \subsection{Data set}
% \label{sec:data-set}

% In \textcite{borchersEfford2008},
% the authors illustrate a potential approach 
% for SCR model selection. The data set used consists of captures of birds
% (red-eyed vireos) from a mist-netting survey on the Patuxent Research Refuge 
% in Maryland, USA, conducted by C.S. Robbins from 1961 to 1972. A variety of SCR
% models are fitted and the final model is selected via AIC. Considering
% a range of parameter dependencies, a candidate set of 
% fifteen selected models is used in \textcite{borchersEfford2008}.
% Working with a similar data set, we use the same parameter
% dependencies. Considering all possible combinations of these
% dependencies allows construction of a comprehensive candidate set of

% forty five models.   

% The `ovenbird' data set, available via the `secr' R package, contains adult captures from a
% mist-netting survey of ovenbirds conducted from 2005 to 2009 at the Patuxent
% Research Refuge by D. K. Dawson and M. G. Efford. As
% described in the documentation, the forest in the area has
% changed little since the previous survey conducted by C.S. Robbins
% \parencite{secr-pkg}. One bird was killed by a predator while in a net
% in 2009 \parencite{secr-pkg}, and for our analysis we remove this loss
% from the data set. We retain any 
% records of individual birds being captured more than once in a
% day. 

% Forty four mist nets were passively operated for nine or
% ten non-consecutive days in late May and June of each year. In the
% context of SCR, populations separated by time (e.g. 
% years) are referred to as `sessions'
% \parencite{effordBook2009}. Therefore, our 
% data set contains five sessions, each with nine or ten sampling occasions. As
% is standard when fitting multi-session SCR models, we each session as
% if from an independent closed population -- that is, we do not
% consider any changes in population size due to births, deaths or migration
% \parencite{effordBook2009}. 

% To construct an accurate picture of the benefits afforded by model
% selection via score tests and AD tools, we implement model selection
% via AIC values alongside our proposed approach. We compare a new, more efficient,
% approach to model selection with a widely-used traditional approach
% using an extensive candidate set of models.

% \subsection{Candidate set}
% \label{sec:candidate-set}

% Please refer to Section \ref{sec:overview-candidate-models} of the Appendix
% for a brief overview of the parametrisation of the candidate SCR
% models. In summary,  
% we have three parameters: $D$, $g_0$ and $\sigma$. $D$
% describes animal density, while $g_0$ and $\sigma$ describe 
% the probability of being captured. For each of these parameters, we
% consider the dependencies presented in
% \textcite{borchersEfford2008}, also shown in Table
% \ref{table:dependencies}, where we specify these dependencies using the
% parlance of the `secr' R package. Considering all possible combinations of
% these dependencies produces the forty five candidate models.

% \begin{table}[h!]
%   \centering
% \begin{tabular} { |c|c|c|c|  }
%   \hline
%   \textbf{$D$ dependencies} & \textbf{$\sigma$ dependencies} &
%                                              \multicolumn{2}{|c|}{\textbf{$g_0$
%                                                                dependencies}} 
%   \\
%   \hline
%   $D \sim 1$ & $\sigma \sim 1$  & $g_0 \sim 1$ & $g_0 \sim \text{b}$ \\
%   \hline
%   $D \sim \text{session}$ & $\sigma \sim \text{session}$ & $g_0 \sim
%                                                            \text{session}$
%              &  $g_0 \sim \text{session} + \text{b}$ \\
%   \hline
%   $D \sim \text{Session}$ & $\sigma \sim \text{h2}$ & $g_0 \sim
%                                                       \text{h2}$ & \\
%   \hline
% \end{tabular}
% \caption{All parameter dependencies considered in our candidate set.}
% \label{table:dependencies}
% \end{table}

% In Table \ref{table:dependencies} (see \textcite{secr-book} for further detail):
% \begin{itemize}
% \item `$\sim 1$' represents a parameter being held pooled across
%   sessions. For example, with $D \sim 1$, we have only one
%   estimate of $D$ and use it for all sessions;
% \item `$\sim \text{session}$' means a parameter has session-specific
%   levels. If we specify $D \sim \text{session}$, we have a separate estimate
%   of $D$ for each session;
% \item `$\sim \text{Session}$' represents a trend across
%   sessions. With `$D \sim \text{Session}$', we  have $D(y) =
%   exp(\beta_0 + \beta_1 y)$, 
%   where $y$ is the session number \parencite{borchersEfford2008}. In
%   our case, $y=1,...,5$;
% \item `$\sim \text{h2}$' denotes a
%   two-class finite mixture model. For example, with `$\sigma \sim \text{h2}$', we assume
%   there are two groups in the population: Group 1 uses $\sigma_1$,
%   while Group 2 uses $\sigma_2$. An 
%   individual falls in Group 1 with probability $p$ and in Group 2
%   with probability $(1-p)$. These estimates are pooled across sessions; 
% \item `$\sim \text{b}$' represents a trap response. With $g_0 \sim
%   \text{b}$, we have two estimates of $g_0$: one for individuals
%   captured on previously in a session, and one for individuals
%   not yet been captured in a session. These estimates are pooled
%   across sessions;
% \item `$\sim \text{session} + \text{b}$' means we have
%   session-specific levels and a trap response -- e.g. with $g_0 \sim \text{session} +
%   \text{b}$, we fit a model with two separate estimates of $g_0$ for
%   each session: one for previously captured individuals, and one for
%   individuals not caught yet. 
% \end{itemize}

% See \textbf{Table (no)} of the Results for all models
% considered. For similarly extensive candidate sets,
% using score tests via the framework of \textcite{catchpoleMorgan1996}
% requires fitting much fewer models than traditional techniques of comparison
% such as LRTs and AIC values. Combining this advantage with the newfound
% convenience provided by AD tools significantly
% enhances the accessibility of model selection.

% Evaluating an SCR likelihood requires a product of 
% two-dimensional integrals, where each integral is an infinite sum over
% the model's random effects \parencite{borchersEfford2008}. Fitting
% every model in a candidate set such as ours requires significant
% resources. Despite this, the ease of calculating AIC values means they
% are commonly used for model selection
% \parencite{efford-schofield}. Although popular models for estimating   
% animal abundance, the complexity of SCR likelihoods has meant derivatives
% and therefore score tests were previously elusive. The benefits afforded by reducing  
% the number of models to be fitted, while harnessing AD tools for
% convenient score tests, are substantial.

% \subsection{Issues with nesting}
% \label{sec:nesting-issues}

% \textbf{Note: in case of mixture models score test issues, McLachlan and Peel
%   book (see Zotero) is a great reference!}

% When working with comprehensive candidate sets, issues with nesting
% may arise. There may be models in the candidate set that do not 
% fit into the levels described above. We extend the framework of
% \textcite{catchpoleMorgan1996} to accommodate this.

% For models to correctly fit into the levels of our chosen model
% selection framework, they must be nested appropriately for score
% tests. There must be only one choice of $\thetaOneTilde$. This ensures
% any changes in $\thetaOneTilde$ cause a change in the $H_1$
% log-likelihood, resulting in the required curvature at
% $\thetaOneTilde$ for the score test statistic to inform how far
% $\thetaOneTilde$ is from the peak.

% This may be violated if our candidate set contains
% models with different parameters. Where issues arise, we propose extending
% the framework of \textcite{catchpoleMorgan1996} by separating the
% candidate set into groups of models with the same parameters and
% implementing the model selection framework on each group. The chosen
% models can be compared using AIC values to select the 
% final model. Being a well-established and straightforward method
% for comparing a few models, using AIC values facilitates our extension
% of the chosen framework, allowing practitioners to harness the
% convenience of score tests even when issues with nesting arise.

% For example, as shown in Table \textbf{(table no)} in the Results, our
% candidate set includes models with and without mixture components. Specifically,
% Model \textbf{(1st model no)} 
% has a mixture component for $\sigma$, while holding $D$ and $g_0$ constant
% for all animals across all sessions. Recall, this means the model
% assumes the ovenbirds fall into one of two groups -- Group 1 uses
% $\sigma_1$ and Group 2 uses $\sigma_2$. The probability of falling
% into either group is $p$ and $(1-p)$, respectively. In contrast, Model
% \textbf{(2nd model no)} holds all three parameters constant for all birds and
% sessions. If we set up a score test to compare these two models there
% is no single value of $\thetaOneTilde$ that, when
% plugged into the $H_1$ model (Model \textbf{(1st model no)}), returns
% the $H_0$ model. \textit{(Elaborate on this -- multiple choices due to
%   any $p$ being appropriate!)} 

% \textit{-- In our example, have models with different mixture
% dependencies. Means do not fit nesting structure required for score
% tests (use explanation below).}

% \textit{-- Doesn't always apply when added parameters, just when added
% parameters result in identifiability issue above -- e.g. behavioural SCR model, can just set
% behavioural effect to 0. }

% % \textit{DO results table then come back to this session. Will clarify how to
% % label models with explanations of dependencies in Appendix... just
% % identify by name? And say something like `we use a mixture model for
% % the parameter labelled $g_0$ that means we don't have the required
% % identifiability when...' and so on?}

% % % We apply the framework of \textcite{catchpoleMorgan1996}, extending it
% % % to a comprehensive candidate set in which all models do not fit
% % % neatly
% % % into a nested structure appropriate for score tests.

% % % For two models to be nested appropriately for score tests, it must be
% % % possible to obtain the simpler model from the more complex one by
% % % placing fixed constraints on any additional parameters. Any small changes in
% % % the values of these constrained parameters must change the
% % % likelihood. This ensures the $H_1$ log-likelihood has curvature at the
% % % $H_0$ estimates, allowing us to calculate the required gradients.

% % % From ChatGPT (for own understanding):
% % % For a score test, we only need the likelihood to have curvature in the
% % % directions of the parameters being tested under H_0 (psi), because the
% % % test measures the sensitivity of the likelihood to changes in those
% % % parameters. Curvature in other parameters is not required — it does
% % % not affect the calculation of the score statistic or its chi-square
% % % distribution.
% % % The score test looks only at the change in likelihood if we move the
% % % additional parameters psi away from their null values (constrained values under
% % % H_0).

% % \textbf{(Perhaps refine with lambda and psi notation, if used in
% %   section above?)}

% % For two nested models to be appropriate for a score test, it must be
% % possible to obtain the simpler model (the $H_0$ model) from the more
% % complex one (the $H_1$ model) via fixed constraints on any additional
% % parameters, $\boldsymbol{\psi}$. All parameters in
% % $\boldsymbol{\psi}$ must be identifiable 
% % in the context of the $H_0$ model -- i.e. when identifying the point on the
% % $H_1$ likelihood surface equivalent to the $H_0$ MLE vector,
% % there must be only one choice. When this holds, any small deviations
% % from this point cause a
% % change in the likelihood, ensuring the $H_1$ log-likelihood 
% % at the $H_0$ MLE vector has the required curvature in the direction of
% % all parameters in $\boldsymbol{\psi}$. Without this curvature, it is no longer clear how
% % far the $H_0$ MLE vector is from the peak of the $H_1$ log-likelihood
% % -- that is, it is unclear how much the additional parameters
% % in $\boldsymbol{\psi}$ improve fit. Any 
% % score statistics are degenerate and the 
% % asymptotic Chi-squared distribution gives unusable
% % p-values. 

% % In our example, issues arise as we consider models that allow for individual
% % heterogeneity. As mentioned in Section \ref{sec:candidate-models}, our candidate set
% % contains models that have a two-class finite mixture dependency for $g_0$ and/or
% % $\sigma$. These models allow individuals in the population to belong to
% % one of two classes, where each class is defined by its corresponding estimate of $g_0$
% % and/or $\sigma$ \textbf{(cite -- SECR book)}. The probability of
% % belonging to either class is estimated by the model. We refer to this
% % probability estimate as $\hat{p}$.

% % % Due to their nature, when conducting score tests, models with mixture
% % % models can only be compared to models with matching mixture
% % % components.  
% % % Consider the example where
% % % our $H_0$ model has no mixture component 
% % % and our $H_1$ model has a mixture component for $g_0$.


% \section{Appendix}
% \label{sec:appendix}

% \subsection{SCR overview and candidate set}
% \label{sec:overview-candidate-models}

% Using records of when and where individuals were captured, SCR models
% allow for estimation of animal density and distribution
% \parencite{borchersEfford2008}. As described in \textcite{secr-book},
% there are two parts to an SCR model: the state model and the
% observation model. The state model describes the distribution of the animal population,
% the `biological reality we wish to describe'. The observation model
% describes the sampling process we use to understand this
% population. Selecting appropriate state and observation models that accurately
% reflect the population and survey enhance the validity of
% inference. For understanding, we briefly describe our selections below. Please refer to
% texts such as \textcite{secr-book} for further details.

% Specifying the state model means choosing a spatial point process to
% describe the distribution of animals in the survey area
% \parencite{borchersFewster2016}. It is common to select a spatial
% Poisson process \parencite{secr-book}. Specifically, following the
% example of \parencite{borchersEfford2008}, we use a
% homogeneous Poisson process, so animals are assumed to be
% distributed uniformly across the survey area. Using the state model,
% we estimate the parameter $D$, animal density in the area of interest.

% To specify the observation process, we select a detection function for
% our SCR models. This describes how an animal's probability of detection
% changes as a function of the distance between their activity centre and a
% detector. An animal's activity centre is the centre of their home
% range \parencite{secr-book}.

% The choice of detection function dictates the `detection' parameters
% to be estimated. Although \textcite{borchersEfford2008} consider
% SCR models with half-normal and hazard rate detection functions, we only consider models with a
% half-normal detection function. Therefore, our detection
% parameters are $g_0$ and $\sigma$. Hazard rate detection
% functions are seldom used and generally not recommended due to their
% long tail \parencite{secr-pkg}.

% Refer to texts such as \textcite{secr-book} for further details on
% state models and observation models. 

% When fitting a SCR model, along with the
% state/observation models, it is standard to specify any dependencies for $D$
% and the detection parameters. Following the example set by
% \textcite{borchersEfford2008}, we 
% consider models with the dependencies shown in Table
% \ref{table:dependencies} (see Section \ref{sec:candidate-set}). 



\printbibliography

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
